Towards a classification framework for social machines

### 2013-02-26 latex/paper_new.tex
Correct use of dashes and removed weird double spaces throughout

\section{Social machines: an introduction}
Once upon a time `machines' were programmed by programmers and used by users. The success of the Web has changed this relationship: we now see configurations of people interacting with content and with each other, typified by social Web sites. Rather than drawing a line through such Web-based systems to separate the human and digital parts (as computer science has traditionally done), we can now draw a line around them  and  treat  each  such  compound  as  a  `social  machine'  -  a  machine in which the two aspects are seamlessly interwoven. This crucial transition in thinking acknowledges the reality of current socio-technical systems, and is essential to underpin any understanding of the science and engineering of their future development towards pervasive ecosystems of co-evolving social machines.

@@

\subsection{Methodology}
\label{sec:methodology}
The classification framework presented in this paper was created through knowledge elicitation ~\cite{knowledgeelicitation}. In particular we used the {\it repertory grid} elicitation
technique~\cite{kelly} in order to derive an initial set of {\it elements}, which represent instances of social machines, and {\it constructs}, which capture their most important characteristics.
In repertory grid elicitation a software tool is used to ask users to describe constructs
that differentiate between elements - for example,  prompting the user to create a construct that
differentiates between {\it GalaxyZoo}\footnote{In GalaxyZoo people classify galaxies with collective performance as good as professional astronomers, see \url{http://www.galaxyzoo.org}.} and {\it Facebook}, as two prominent example of social machines our classification framework should apply to. The user describes the opposing poles of the construct - in this case, the user may decide on `For Science' and `For socialising' in order to capture the core distinction between the two system. The user then rates every element with value from $1$ to $5$ on this construct, where $1$ represents an element that is purely
`For Science', and $5$ one that mainly serves  `socialising' purposes. Elements can also be rated with values between the poles if they may correspond to both or neither. The repertory grid software then identifies elements that are the least differentiated when requesting new constructs, in an attempt to elicit as much knowledge as possible from the user. We applied this technique to create an initial set of examples of social machines and to classify them according to specific features. To do so, we asked then computer science researchers familiar with the field to create their own repertory grids, and populate the elements from their own knowledge, and created the constructs using the standard repertory grid elicitation technique described above. This exercise led to $10$ grids, the union of which comprised a total of $56$ unique elements (social machines)
and $117$ different constructs (classifying factors). As the aim of this initial phase was to understand how people perceive the notion of social machine and their most distinctive properties, we allowed the participants to choose the social machines they are familiar with and describe them in their own terms.

@@

\subsubsection{Participants and roles}
This set of constructs pertain to the human participants of social machines and the way(s) participants are organized within them. One construct identifies whether participants constitute members of the general public, or are of a specific demographic, external group or organization, occupation, or of a particular expertise.  Two constructs, \emph{autonomy} and \emph{anonymity} pertain to the degree to which interaction is constrained among participants, and to the degree to which participants' identities are used and disclosed within the machine. Additional two constructs pertain to the roles: first, as to whether multiple roles are pre-differentiated by the machine or
whether all participants essentially initially assume the same role, second whether roles (pre-defined or emergent) are hierarchical.

\subsubsection{Motivation and incentives}
Related to the social structure  motivation that sustains
continued participation in these systems.  We identify six different
kinds of common sources of intrinsic motivation; first, that
participation is fun, second that it accomplishes an activity that the
participant enjoys or wanted to get done, third, that it satisfies the
desire to gain or share knowledge, or, fourth, or to be social.  The
last two pertain to philanthropy, whether participation is seen to
benefit a particular group of disadvantaged people or individuals who
need assistance, or, finally, if participation is beneficial to
society as a whole.  Finally, we ask whether extrinsic motivational
factors also contributed to sustained participation - such factors
could include money reward, status, recognition and so on \cite{kuznetsov2006motivations,lakhani2003hackers,tagging}.
%Peer-to-peer filesharing~\cite{p2p}; Tagging~\cite{tagging}; and Virtual
%Communities~\cite{ardichvili2003motivation,Moore:2007:UMM:1235000.1235035}.

@@

% \subsubsection{Users and Roles}
% {\bf The roles of the participants and the social network they form}
% \subsubsection{Participation (How users interact with the Social Machine)}
%\subsubsection{Participation - \emph{How and when user participate}}
%\subsubsection{Problems Solved and Benefit Derived}
%The result is what we call `participation' constructs. These constructs arise from use of a
%social machine, and are not necessarily predictable; they include:
%
%\begin{enumerate}
%\item {\bf Roles emerging in the social machine}
%   \newline In addition to designed and implemented roles such as {\it moderator} and {\it administrator}, emerging roles include motivations of the participant, such as {\it trolls}, {\it spammers}, {\it domain experts}, {\it social nexus}, and {\it re-blogger}.
%\item {\bf Quality of contributions}
%    \newline For any social machine, the quality of contributions will vary. While some
%    issues with quality can be predicted and protected against (e.g., through
%    multi-user result agreement on Mechanical Turk~\cite{ipeirotis2010quality}), some
%    particular issues with quality cannot always be predicted ahead of time.
%\item {\bf Content of contributions}
%    \newline While a system operator may design their social machine to generate specific
%    types of content, participants may subvert the system in order to generate other types,
%    or to concentrate on areas that were unforeseen at the time of launch, but are now seen
%    as useful to the operators. It is also possible that multiple social machines can be
%    combined/linked by users to further expand the types of contributions they produce.
%\item {\bf Global / cultural reach}
%    \newline A social machine may naturally focus on a narrow geographic area, for example
%    an Ushahidi-based social machine for local elections~\cite{meier2008crisis} or
%    natural disasters~\cite{morrow2011independent}. Such machines are unlikely to be
%    re-purposed by participants outside of that area. However, some machines have been
%    created for a single cultural/global area, but have potential global relevance that was
%    unforeseen early on. An example is the fall-off of English-language users of the
%    social network ``Orkut,'' which gained significant usage elsewhere, particularly in
%    Brazil.
%\item {\bf Task reach --- what do people use it for?}
%    \newline In addition to breaking cultural barriers, user have also shown that they can
%    use social machines for tasks that were not set up by the system operators. For example
%    the identification of ``Green Pea Galaxies'' using the GalaxyZoo social machine for
%    astronomy~\cite{greenpea}. Users were not originally asked to identify these types, but
%    a number of users identified them, and the software was modified to include them,
%    resulting in their discovery as a new type of galaxy.
%\item {\bf Active/Passive participation roles}
%    \newline When it comes to participation in social machines, there are levels of
%    participation that can be classified as being ``active'' and being ``passive.'' These
%    roles may not be particularly designed into systems, but may be exhibited over time.
%    Particularly noteworthy examples can be seen in news and link aggregation social
%    machines such as Reddit and Digg, where users can both submit links and vote on links.
%    The frequency of each of these activities can vary wildly, with a small
%    minority of users controlling the links that rise to the top~\cite{digg}.
%\end{enumerate}

@@

We envisage three key use cases for our framework. First and foremost the framework forms the basis for a coherent and comprehensive description and classification of social machines. The knowledge elicitation technique referred to in this paper is just one of the many tools which can assist this exercise. In addition, it enables direct comparisons between systems, and provides a platform for terminologically consistent discussions about the field as a whole and specific instantiations. Finally, we also think that with more and more classification data becoming available, for instance, also through functionality that automatically monitors and computes the values of the some of the constructs introduced earlier, researchers will wish to apply prediction techniques (such as SVMs) to social machines, their behavior, and impact.

\section{Future work: building a social machine around the classification framework}
In this paper we presented a first outline of a possible organization structure for the design space of social machines. We introduced the key features of the notion of social machines, and explained how this notion differs from related areas such as collective intelligence, crowdsourcing and human computing. We then discussed the basic constructs of a classification framework created through the repertory grid elicitation technique. In its current version the framework consists of $31$ constructs clustered according to the main components of social machines: social and machine-driven services, and the interactions between them. Our future work will mostly be dedicated to the revision of this initial list of constructs. Most notably, we would like to add a number of constructs to emphasize the importance of the interaction aspect in the theory, engineering, and operation of social machines. This includes constructs related to the human-computer workflows supported (e.g., human intelligence replacing automatic algorithms as in GWAPs vs the different bots available on Wikipedia complementing collaborative manual editing), the way inputs from the crowd are validated, aggregated and turned into a useful outcome, as well as issues related collaboration and coordination, and interface and communication design. In addition, we plan to engage with the broader researcher community by setting up a social machine to facilitate discussions and feedback collection. One specific idea that is currently under development is to build a microtask-like environment, including specific game elements, in which participants are asked to provide answers to atomic challenges that rate and compare a pair of social machine instances according to a construct of our framework.

>>>>

\section{Social machines: an introduction}
Once upon a time `machines' were programmed by programmers and used by users. The success of the Web has changed this relationship: we now see configurations of people interacting with content and with each other, typified by social Web sites. Rather than drawing a line through such Web-based systems to separate the human and digital parts (as computer science has traditionally done), we can now draw a line around them and treat each such compound as a `social machine' --- a machine in which the two aspects are seamlessly interwoven. This crucial transition in thinking acknowledges the reality of current socio-technical systems, and is essential to underpin any understanding of the science and engineering of their future development towards pervasive ecosystems of co-evolving social machines.

@@

\subsection{Methodology}
\label{sec:methodology}
The classification framework presented in this paper was created through knowledge elicitation ~\cite{knowledgeelicitation}. In particular we used the {\it repertory grid} elicitation
technique~\cite{kelly} in order to derive an initial set of {\it elements}, which represent instances of social machines, and {\it constructs}, which capture their most important characteristics.
In repertory grid elicitation a software tool is used to ask users to describe constructs
that differentiate between elements -- for example, prompting the user to create a construct that
differentiates between {\it GalaxyZoo}\footnote{In GalaxyZoo people classify galaxies with collective performance as good as professional astronomers, see \url{http://www.galaxyzoo.org}.} and {\it Facebook}, as two prominent example of social machines our classification framework should apply to. The user describes the opposing poles of the construct -- in this case, the user may decide on `For Science' and `For socialising' in order to capture the core distinction between the two system. The user then rates every element with value from $1$ to $5$ on this construct, where $1$ represents an element that is purely
`For Science', and $5$ one that mainly serves `socialising' purposes. Elements can also be rated with values between the poles if they may correspond to both or neither. The repertory grid software then identifies elements that are the least differentiated when requesting new constructs, in an attempt to elicit as much knowledge as possible from the user. We applied this technique to create an initial set of examples of social machines and to classify them according to specific features. To do so, we asked then computer science researchers familiar with the field to create their own repertory grids, and populate the elements from their own knowledge, and created the constructs using the standard repertory grid elicitation technique described above. This exercise led to $10$ grids, the union of which comprised a total of $56$ unique elements (social machines)
and $117$ different constructs (classifying factors). As the aim of this initial phase was to understand how people perceive the notion of social machine and their most distinctive properties, we allowed the participants to choose the social machines they are familiar with and describe them in their own terms.

@@

\subsubsection{Participants and roles}
This set of constructs pertain to the human participants of social machines and the way(s) participants are organised within them. One construct identifies whether participants constitute members of the general public, or are of a specific demographic, external group or organisation, occupation, or of a particular expertise. Two constructs, \emph{autonomy} and \emph{anonymity} pertain to the degree to which interaction is constrained among participants, and to the degree to which participants' identities are used and disclosed within the machine. Additional two constructs pertain to the roles: first, as to whether multiple roles are pre-differentiated by the machine or
whether all participants essentially initially assume the same role, second whether roles (pre-defined or emergent) are hierarchical.

\subsubsection{Motivation and incentives}
Related to the social structure motivation that sustains
continued participation in these systems. We identify six different
kinds of common sources of intrinsic motivation; first, that
participation is fun, second that it accomplishes an activity that the
participant enjoys or wanted to get done, third, that it satisfies the
desire to gain or share knowledge, or, fourth, or to be social. The
last two pertain to philanthropy, whether participation is seen to
benefit a particular group of disadvantaged people or individuals who
need assistance, or, finally, if participation is beneficial to
society as a whole. Finally, we ask whether extrinsic motivational
factors also contributed to sustained participation -- such factors
could include money reward, status, recognition and so on \cite{kuznetsov2006motivations,lakhani2003hackers,tagging}.
%Peer-to-peer filesharing~\cite{p2p}; Tagging~\cite{tagging}; and Virtual
%Communities~\cite{ardichvili2003motivation,Moore:2007:UMM:1235000.1235035}.

@@

% \subsubsection{Users and Roles}
% {\bf The roles of the participants and the social network they form}
% \subsubsection{Participation (How users interact with the Social Machine)}
%\subsubsection{Participation -- \emph{How and when user participate}}
%\subsubsection{Problems Solved and Benefit Derived}
%The result is what we call `participation' constructs. These constructs arise from use of a
%social machine, and are not necessarily predictable; they include:
%
%\begin{enumerate}
%\item {\bf Roles emerging in the social machine}
%  \newline In addition to designed and implemented roles such as {\it moderator} and {\it administrator}, emerging roles include motivations of the participant, such as {\it trolls}, {\it spammers}, {\it domain experts}, {\it social nexus}, and {\it re-blogger}.
%\item {\bf Quality of contributions}
%  \newline For any social machine, the quality of contributions will vary. While some
%    issues with quality can be predicted and protected against (e.g., through
%    multi-user result agreement on Mechanical Turk~\cite{ipeirotis2010quality}), some
%    particular issues with quality cannot always be predicted ahead of time.
%\item {\bf Content of contributions}
%    \newline While a system operator may design their social machine to generate specific
%    types of content, participants may subvert the system in order to generate other types,
%    or to concentrate on areas that were unforeseen at the time of launch, but are now seen
%    as useful to the operators. It is also possible that multiple social machines can be
%    combined/linked by users to further expand the types of contributions they produce.
%\item {\bf Global / cultural reach}
%    \newline A social machine may naturally focus on a narrow geographic area, for example
%    an Ushahidi-based social machine for local elections~\cite{meier2008crisis} or
%    natural disasters~\cite{morrow2011independent}. Such machines are unlikely to be
%    re-purposed by participants outside of that area. However, some machines have been
%    created for a single cultural/global area, but have potential global relevance that was
%    unforeseen early on. An example is the fall-off of English-language users of the
%    social network ``Orkut,'' which gained significant usage elsewhere, particularly in
%    Brazil.
%\item {\bf Task reach --- what do people use it for?}
%    \newline In addition to breaking cultural barriers, user have also shown that they can
%    use social machines for tasks that were not set up by the system operators. For example
%    the identification of ``Green Pea Galaxies'' using the GalaxyZoo social machine for
%    astronomy~\cite{greenpea}. Users were not originally asked to identify these types, but
%    a number of users identified them, and the software was modified to include them,
%    resulting in their discovery as a new type of galaxy.
%\item {\bf Active/Passive participation roles}
%    \newline When it comes to participation in social machines, there are levels of
%    participation that can be classified as being ``active'' and being ``passive.'' These
%    roles may not be particularly designed into systems, but may be exhibited over time.
%    Particularly noteworthy examples can be seen in news and link aggregation social
%    machines such as Reddit and Digg, where users can both submit links and vote on links.
%    The frequency of each of these activities can vary wildly, with a small
%    minority of users controlling the links that rise to the top~\cite{digg}.
%\end{enumerate}

@@

We envisage three key use cases for our framework. First and foremost the framework forms the basis for a coherent and comprehensive description and classification of social machines. The knowledge elicitation technique referred to in this paper is just one of the many tools which can assist this exercise. In addition, it enables direct comparisons between systems, and provides a platform for terminologically consistent discussions about the field as a whole and specific instantiations. Finally, we also think that with more and more classification data becoming available, for instance, also through functionality that automatically monitors and computes the values of the some of the constructs introduced earlier, researchers will wish to apply prediction techniques (such as SVMs) to social machines, their behaviour, and impact.

\section{Future work: building a social machine around the classification framework}
In this paper we presented a first outline of a possible organisation structure for the design space of social machines. We introduced the key features of the notion of social machines, and explained how this notion differs from related areas such as collective intelligence, crowdsourcing and human computing. We then discussed the basic constructs of a classification framework created through the repertory grid elicitation technique. In its current version the framework consists of $31$ constructs clustered according to the main components of social machines: social and machine-driven services, and the interactions between them. Our future work will mostly be dedicated to the revision of this initial list of constructs. Most notably, we would like to add a number of constructs to emphasize the importance of the interaction aspect in the theory, engineering, and operation of social machines. This includes constructs related to the human-computer workflows supported (e.g., human intelligence replacing automatic algorithms as in GWAPs vs the different bots available on Wikipedia complementing collaborative manual editing), the way inputs from the crowd are validated, aggregated and turned into a useful outcome, as well as issues related collaboration and coordination, and interface and communication design. In addition, we plan to engage with the broader researcher community by setting up a social machine to facilitate discussions and feedback collection. One specific idea that is currently under development is to build a microtask-like environment, including specific game elements, in which participants are asked to provide answers to atomic challenges that rate and compare a pair of social machine instances according to a construct of our framework.
